\section{Parallelism}
Writing parallel programs that operate on a large amount of data can be difficult.
To achieve the best performance on a multi-processor machine, programs must be written using low-level synchronisation techniques such as threads and locking.
These methods are error prone, and improper use can lead to deadlocks and Heisenbugs: such errors are inherently hard to debug.

There are several techniques for making parallelism easier to take advantage of,
each with their own advantages and disadvantages.
We discuss three techniques:

\begin{description}
\item[Flat   data parallelism]
is able to achieve speeds comparable to C with advanced fusion techniques,
but only supports regular structures, such as rectangular and cubic arrays.

\item[Nested data parallelism]
is similar to flat data parallelism,
but supports parallel operations over tree-like structures
while keeping the load balanced over processors.

\item[Feedback directed automatic parallelism]
is a different beast: it uses profiling information from running the program
to find the best places to introduce parallelism. 
\end{description}


\subsection{Flat data parallelism}
Flat data parallelism such as Repa\cite{keller2010regular} and Accelerate\cite{mcdonell2013optimising}
is a restricted form of data parallelism that disallows nested parallelism: only the top-most computation is performed in parallel.
To ensure a balanced load across parallel computations, only regular structures such as rectangular arrays can be used.

Accelerate is an \emph{embedded domain-specific language} in Haskell, specifically for running parallel computations on graphics processors.
As a restricted domain-specific language, it targets graphics processors\cite{chakravarty2011accelerating} and is able to use fusion\cite{mcdonell2013optimising}.

Repa is a Haskell library with combinator support for regular parallel array computations\cite{keller2010regular}.
It uses type indices to guide fusion\cite{lippmeier2012guiding} and fancy footwork to support efficient stencil convolution\cite{lippmeier2011efficient}.


\subsection{Nested data parallelism}
Data parallel Haskell\cite{chakravarty2007data}, \nesl\cite{blelloch1995nesl}.


\subsubsection{Flattening}

\rev{Enlarging the Scope of Vector-Based Computations: Extending Fortran 90 by Nested Data Parallelism}{au1997enlarging}
Vector computers are good for flat data parallelism, which requires regular data, and so languages like Fortran only support rectangular array parallelism.
Nested data parallelism can be converted to flat data parallelism with a \emph{flattening} transform.
An extension to Fortran 90 is shown that flattens nested data parallelism, allowing existing optimising Fortran compilers to use the vector machinery.

\rev{Scans as Primitive Parallel Operations}{blelloch1989scans}
If distributed vector machines implemented scan operations on vectors, much time could be saved that would otherwise be spent waiting for memory accesses to return.
Explanation of radix sort example.
Segmented scan operations use a vector of flags the same length as the values. I guess this totally disallows empty segments, which makes sense. True flag indicates the start of a segment. What happens if there is no True flag at the start..?
Clever implementation of Quicksort using segmented scans and iteratively splitting each segment into more subsegments. Using the segment flag representation allows very easy splitting.
It seems quite disingenuous to claim ``constant time for a vector of size n, provided there are n nodes in the machine''. It does deal with this eventually though, in the load-balancing section.
Hardware description of a tree scan.

\rev{Vector Models for Data-Parallel Computing}{blelloch1990vector}
Book covering scan primitives and basic vectorisation/flattening.
Must reread this to understand the flattening they use.

\rev{A Comparison of Sorting Algorithms for the Connection Machine CM-2}{blelloch1991comparison}
Different sorting algorithms are implemented and compared for the connection machine. Irrelevant.

\rev{Implementation of a Portable Nested Data-Parallel Language}{blelloch1993implementation}
Goes through the implementation of \nesl, but only the VCODE (virtual machine) and CVL (parallel primitives library).
Doesn't actually talk about flattening or vectorisation.
Irrelevant.

\rev{\nesl: A Nested Data-Parallel Language}{blelloch1995nesl}
This is more like a user guide / tutorial for programming in \nesl, rather than any information about how it is implemented.

\rev{Programming Parallel Algorithms}{blelloch1996programming}
More example programs in \nesl...


\rev{How Portable is Nested Data Parallelism?}{chakravarty1999portable}
Compilers for nested data parallelism have been implemented for many different architectures (vector, scalar / single, shared or distributed memory)
but previously these compilers have only targeted one architecture with optimisations specific to that architecture.
These different architectures require significantly different optimisation techniques to achieve good performance:
vector machines work well with regular vector operations.
On the other hand RISC processors, due to their limited registers,
perform better on complex operations on a small set of scalars.

In this paper, flattening is combined with fusion, to remove intermediate allocations and achieve better cache performance.
The flattening transform remains the same for all architectures, but fusion is tuned specific to each architecture.

A potential problem with fusion, however, is that it may in fact reduce parallelism. 
To solve this, another transform is added that partitions the program into local and global operations.
By fusing only local operations across no global boundaries, no parallelism will be lost.

The code generation must also be tuned specific to each architecture; despite targeting C,
care must still be taken to ensure loops map to vector operations on vector machines.


\rev{More Types for Nested Data Parallel Programming}{chakravarty2000more}
Blelloch \& Sabot's original flattening transformation didn't support general sum types, recursive types, or higher-order functions. Recursive types have been shown in \cite{keller1998flattening}.
This generalises flattening to support the full range of types in Haskell.
By formalising the flattening transformation in a lambda calculus they are able to easier express separate compilation.

\rev{Nepal - Nested Data-Parallelism in Haskell}{chakravarty2001nepal}
\rev{An Approach to Fast Arrays in Haskell}{chakravarty2003approach}
\rev{Data Parallel Haskell: a status report}{chakravarty2007data}
\rev{Partial Vectorisation of Haskell Programs}{chakravarty2008partial}

\rev{Harnessing the Multicores: Nested Data Parallelism in Haskell}{jones2008harnessing}


\rev{Flattening trees}{keller1998flattening}
Problem: \nesl etc flattening transform only supports homogenous vectors, so trees are awkward to program.
The flattening transform is extended to allow user-defined recursive types.

\rev{On the distributed implementation of aggregate data structures by program transformation}{keller1999distributed}
Separating local from global computations, so that fusion can be applied to only local computations.
--REREAD


\rev{Costing nested array codes}{lechtchinsky2002costing}
\rev{Higher order flattening}{leshchinskiy2006higher}

\rev{Work efficient higher-order vectorisation}{lippmeier2012work}


\subsubsection{Vectorisation}

\rev{Higher-order nested data parallelism: semantics and implementation}{leshchinskiy2005higher}
Formalisation of closure conversion and flattening. 

In the standard lambda calculus, data and code are somewhat inseparable with partial application capturing data.
This means that the flattening transform is not able to operate on partial applications or higher-order functions.
Closure conversion makes the capturing of data explicit, allowing the flattening transform to operate on captured data.

The flattening transform converts nested data parallelism into flat data parallelism.
This is achieved by storing nested arrays as a flat data array and a separate segment descriptor describing the layout of the nesting.
Each function is then `lifted' to a parallel computation, producing a new function equivalent to $\mbox{map} f$.

Nested maps on these nested arrays can then be implemented by simply mapping over the data array and reusing the original segment descriptor,
since the structure of the nesting has not changed.

As with all optimisations, conversion from nested data parallelism to flat data parallelism should be semantics preserving:
the original program's meaning must remain unchanged.
This is complicated by the fact that Haskell is a lazy language, whereas, for efficiency, strict unboxed arrays are used to represent data.
Several tricks are used to preserve laziness while still using strict unboxed arrays.


\subsubsection{Vectorisation avoidance}

\rev{Vectorisation avoidance}{keller2012vectorisation}
Flattening converts scalar operations into array operations, even for intermediate scalars.
This means that operations that once were simple additions on registers must now work on arrays in memory.
Array fusion can often remove these intermediate structures, but fusion relies heavily on other optimisations such as inlining,
which may not always occur due to sharing. Excessive inlining can also cause code blowup in the compilation process,
leading to long compile times.

Instead of vectorising an entire program and then relying on fusion to remove intermediate arrays,
it is shown that identifying scalar operations and not vectorising them is less fragile.

This is achieved in three steps:
\begin{enumerate}
\item labelling: each subexpression is classified as parallel, scalar, etc
\item encapsulation: the labelled syntax tree is traversed and maximal sequential subexpressions are lifted.
The lifted lambdas are given a new label, encapsulated
\item vectorisation: nested data parallelism is removed by flattening, but any expressions labelled 'encapsulated' are not vectorised
\end{enumerate}

Pairs of Scalars are not treated.
There is a problem of load imbalance when treating conditionals and recursive functions.
If one branch of an \verb/if/ is significantly expensive, it may turn out that when this is executed over a large vector,
one core has to do all the expensive branches and the others only have simple branches.
One potential solution is to randomise the input vector before balancing it, to make an even distribution more likely.




\subsection{Feedback directed automatic parallelism}
Automatic feedback-directed implicit parallelism in Mercury \cite{bone2010automatic}.

